MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-15 main/MSstatsPTM_Main.tex
addressed some comments

\usepackage{fullpage}
\usepackage{amsfonts,amsthm}
\usepackage{amsmath}
%\numberwithin{figure}{section} % numbering in each subsection
\numberwithin{table}{section}
\usepackage{setspace}
\usepackage{url}
\usepackage{lscape} %% rotating table
\usepackage{rotating}
\usepackage{authblk}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{xcolor} %\[table,dvipsnames]
%\usepackage{makecell}
%\usepackage{array}

@@

\usepackage{tabulary,multirow,multicol,rotating}
\usepackage[backend=biber,sorting=none]{biblatex}
\addbibresource{bibliography.bib}

@@

@@

\author[1]{Devon~Kohler}
\author[2]{Tsung-Heng~Tsai}
\author[1]{Ting~Huang}
\author[4]{Erik~Verschueren}
\author[3]{Trent~Hinkle}
\author[3]{Meena~Choi*}
\author[1]{Olga~Vitek*}
\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
\affil[2]{Kent State University, Kent, OH, USA}
\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
\affil[4]{ULUA BV, Arendstraat 29, 2018 Antwerp, Belgium}
\affil[*]{Corresponding Authors}

@@

Liquid chromatography coupled with bottom up mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) in samples conditions. Analysis of data from such experiments faces numerous statistical challenges. These include the low abundance of modified proteoforms, the small number of observed peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation. In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results. The framework requires an experimental design, which quantifies, for each sample, both peptides with post-translational modifications and peptides from the same proteins with no modification sites. The proposed framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, by fitting separate linear mixed effects models appropriate for the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments \ov{with different organisms, modification types and data acquisition types} demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches. The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.

@@

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}.

@@

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test \todo{ref}, Analysis of Variance\todo{ref}, or Limma\todo{ref}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites~\cite{Ritchie_15a}. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs \todo{as compared to methods XXX}. In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

@@

%%%% Summary of controlled mixtures
\begin{table}[t!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of bio. &  Data \\
&  & design & Conditions & replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - Missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets.} 
```Dataset'' is the dataset code name.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode. \todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition. But it seems that some descriptions imply the total number of biological samples across conditions. Could you clarify that we use number of biological replicates per condition throughout? (E.g, this helps calculate the degrees of freedom in balanced designs)}
}
\label{tab:dataDescription}
\end{table}

@@

\todo{Additional details on every dataset are available in Supplementary Section 2 (I think it's ok to only mention this once)}

\todo{Could you delete this paragraph, and move the container identifies to a new column in table 1?} The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).

@@

{\bf Simulation design:} The simulation represented an idealistic case. 24 synthetic label-free datasets were generated with different experimental designs and different biological variation. In each dataset, 1000 proteins had 10 unmodified features per protein. Each of the 1000 proteins had one PTM. Each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All the differential PTMs were generated with an expected log fold change of 0.75 between conditions.

@@

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods \todo{XXX - what and how}

@@

\medskip \noindent {\bf Evaluation:} \todo{As above, we evaluated... what and how...}

@@

\medskip \noindent {\bf Experimental Design:} Figure \ref{fig:benchmark-design}(a) overviews the experimental design. Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E. coli} lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were viewed as the global proteome.   Background {\it E. coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling \ov{between Mix 1 and 3, and between Mix 2 and 4}.  50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark. \todo{In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed). } The true log-fold changes \ov{between the relevant components of the relevant mixtures} are summarized in Figure \ref{fig:benchmark-design}(b).  Two replicates mixtures were created per condition.

\medskip \noindent {\bf Data acquisition:} Each mixture was analyzed with KGG enrichment, and without KGG enrichment (i.e., in a global profiling run), with label-free LC-MS/MS. There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Evaluation:} \todo{Here is the place to describe: 'we expect XXX to not change'. Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity). We expect XXX to have fold changes XXX between conditions XXX. We evaluated the accuracy of estimation of fold changes... and the sensitivity of detecting differentially abundant...}  
%
\todo{I think some mixtures are not comparable because the total protein amount is not the same. I believe only in Mix 1 vs 2 and Mix 3 vs 4 are comparable. I do not think it is correct to compare Mix 1 vs 3 and 2 vs 4 directly because they differ in the total amount of the proteome. Or was it normalized based on what we know is the same? If yes, we need to describe in the beginning of 'Evaluation'}  \todo{The unadjusted proteins... estimation... The adjusted proteins... estimation... Positive controls... Negative controls...}

@@

\medskip \noindent {\bf Experimental Design:}  Luchetti et al. \cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter. Uninfected cells were measured at 0 and 6 hours, while cells infected \todo{with what? Not clear about the infection} were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. \todo{Is GSDMD a known protein that will be used for validation?}

\medskip \noindent {\bf Data acquisition:} 11 samples \todo{why 11 and not 12? Which condition didn't have a biological replicate?} were allocated to 1 TMT mixture in an unbalanced group comparison design. The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

@@

\subsection*{Dataset 5: Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}

\medskip \noindent {\bf Experimental Design:} Maculins et al. \cite{Maculins} studied primary murine macrophages infected with {\it Shigella flexneri} ({\it S. flexneri}). The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and infected with {\it S. flexneri}. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).

\medskip \noindent {\bf Data acquisition:} 8 biological samples \todo{total? Or per condition? I am having difficulties understanding the allocation from the description...} were allocated to 2 TMT mixtures in an unbalanced time course design. Each mixture contained two replicates per early and late WT and KO conditions. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run. There was a 90\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Time series comparison:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late. 9 total comparisons were made, namely KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT. Since the dataset was a biological investigation, the true positive modifications were unknown.

@@

\medskip \noindent {\bf Experimental Design:} Cunningham et al. \cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. The experiment profiled ubiquitination sites, and analyzed changes in the modified site abundance. The experiment had four conditions, CCCP, USP30 over expression (USP30 OE), Combo, and Control. Each condition had two biological replicates \todo{were these patients or cell lines?}. The abundance of modified peptides was quantified with label-free LC-MS/MS.

@@

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates per condition. For simplicity, we assume that the experiment has no technical replicates, and each biological replicate corresponds to an LC-MS/MS run.  For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$.  \ov{The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.}

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM site and the corresponding unmodified features of the protein, in a special case of 2 conditions and 2 biological replicates per condition. The number of quantified modified and unmodified features varies across proteins. Some log-intensities may be outliers. Some spectral features can be missing.

\todo{OV: from here on, it is unclear if you introduce the general notation, or talk about our model specifically. If it is our model, it should be in the Results} The population parameter we are interested in estimating is the difference between the abundances of a PTM site in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models. For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$. Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}
However, these population parameters are inherently convoluted with changes in the unmodified protein. To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$. We estimate \todo{OV: our method, or everyone?} the quantities $\hat{\mu}_i$ and $\widehat{SE(\hat{\mu}_i)}$ in the same way as for the PTM site, using their corresponding features and observations and resulting in an estimated mean abundance. These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

@@

\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions. While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

>>>>

%\usepackage{fullpage}
\usepackage[margin=1in,headheight=13.6pt]{geometry}
\usepackage{amsfonts,amsthm}
\usepackage{amsmath}
%\numberwithin{figure}{section} % numbering in each subsection
\numberwithin{table}{section}
\usepackage{setspace}
\usepackage{url}
\usepackage{lscape} %% rotating table
\usepackage{rotating}
\usepackage{authblk}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{xcolor} %\[table,dvipsnames]
%\usepackage{makecell}
%\usepackage{array}

@@

\usepackage{tabulary,multirow,multicol,rotating}
\usepackage[backend=biber, citestyle=numeric, bibstyle=authoryear, sorting=none]{biblatex}
\addbibresource{bibliography.bib}

\makeatletter
\input{numeric.bbx}
\makeatother

@@

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[L]{MSstatsPTM: Statistical relative quantification of PTMs}

@@

\author[1]{Devon~Kohler}
\author[2]{Tsung-Heng~Tsai}
\author[4]{Erik~Verschueren}
\author[1]{Ting~Huang}
\author[3]{Trent~Hinkle}
\author[3]{Lilian~Phu}
\author[3]{Meena~Choi*}
\author[1]{Olga~Vitek*}
\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
\affil[2]{Kent State University, Kent, OH, USA}
\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
\affil[4]{ULUA BV, Arendstraat 29, 2018 Antwerp, Belgium}
\affil[*]{Corresponding Authors}
%\affil[]{MSstatsPTM: Statistical relative quantification of PTMs}

@@

Liquid chromatography coupled with bottom up mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) in samples conditions. Analysis of data from such experiments faces numerous statistical challenges. These include the low abundance of modified proteoforms, the small number of observed peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation. In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results. The framework requires an experimental design, which quantifies, for each sample, both peptides with post-translational modifications and peptides from the same proteins with no modification sites. The proposed framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, by fitting separate linear mixed effects models appropriate for the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with different organisms, modification types and data acquisition types demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches. The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.

@@

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}\cite{Kim2016}. This helps to distinguish between relative site occupancy changes at steady-state protein levels, typical for short time-scale signaling events, and observed relative changes of PTMs as a result of underlying gene expression or protein abundance levels.

@@

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test\cite{Kalpic:2011}, Analysis of Variance\cite{girden:1992}, or Limma\cite{Ritchie_15a}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

@@

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs as compared to existing methods. In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

@@

@@

Additional details on every dataset are available in Supplementary Section 2.

@@

{\bf Simulation design:} The simulation represented an idealistic case. 24 synthetic label-free datasets were generated with different experimental designs and different biological variation. In each dataset, 1,000 proteins had 10 unmodified features per protein. Each of the 1,000 proteins had one PTM. Each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All the differential PTMs were generated with an expected log fold change of 0.75 between conditions.

@@

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to correctly detect which PTMs were simulated with changes between conditions and which were not. We evaluated the methods ability to avoid false positives (i.e. specificity), accurately estimate the fold change between conditions, and analyzed the sensitivity of detecting differentially abundant PTMs. Each method was used to analyze changes in PTM abundance in the presence of confounding with changes in the unmodified protein, and with this confounding removed.

@@

\medskip \noindent {\bf Evaluation:} The methods were evaluated in the same way as above. We evaluated their ability to correctly detect PTM's specificity, fold change estimation, and sensitivity. These statistics were analyzed both in the presence of, and without, confounding with changes in the unmodified protein.

@@

\medskip \noindent {\bf Experimental Design:} Figure \ref{fig:benchmark-design}(a) overviews the experimental design. Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E. coli} lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were viewed as the global proteome. Background {\it E. coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling between Mix 1 and 3, and between Mix 2 and 4.  50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides in a mixed background of lysates from {\it E. coli} and a human cell line. Quantitative changes in protein and site abundance of these 20 human proteins were the target of the benchmark. In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed). The true log-fold changes between the relevant components of the relevant mixtures are summarized in Figure \ref{fig:benchmark-design}(b).  Two replicate mixtures were created per condition.

\medskip \noindent {\bf Data acquisition:} Each mixture was analyzed with KGG enrichment, and without KGG enrichment (i.e., in a global profiling run), with label-free LC-MS/MS. There was a 90.2\% overlap of protein identifications between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Evaluation:} We expect the spike-in peptides to change corresponding to the values in Figure \ref{fig:benchmark-design}(b). These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4. Additionally, the background {\it E. coli} lysate peptides are treated as negative controls and are not expected to change in any comparison. Therefor we evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.

\todo{Here is the place to describe: 'we expect XXX to not change'. Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity). We expect XXX to have fold changes XXX between conditions XXX. We evaluated the accuracy of estimation of fold changes... and the sensitivity of detecting differentially abundant...}  
%
\todo{I think some mixtures are not comparable because the total protein amount is not the same. I believe only in Mix 1 vs 2 and Mix 3 vs 4 are comparable. I do not think it is correct to compare Mix 1 vs 3 and 2 vs 4 directly because they differ in the total amount of the proteome. Or was it normalized based on what we know is the same? If yes, we need to describe in the beginning of 'Evaluation'}  \todo{The unadjusted proteins... estimation... The adjusted proteins... estimation... Positive controls... Negative controls...}

@@

\medskip \noindent {\bf Experimental Design:}  Luchetti et al. \cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter. Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella flexneri} ({\it S. flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. Shigella ubiquitin ligase IpaH7.8 was shown to function as an inhibitor of the protein gasdermin D (GSDMD). GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment in human cells.

\medskip \noindent {\bf Data acquisition:} 11 samples were allocated to 1 TMT mixture in an unbalanced group comparison design. All conditions were allocated two biological replicates except for the Dox1hr condition, which was allocated one replicate. The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

@@

\subsection*{Dataset 5: Mouse - Phosphorylation - 2mix-TMT}
\label{sec:exp_proc_dataset5}

\medskip \noindent {\bf Experimental Design:} Maculins et al. \cite{Maculins} studied primary murine macrophages infected with {\it S. flexneri}. The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and infected with {\it S. flexneri}. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).

\medskip \noindent {\bf Data acquisition:} 22 biological samples were allocated to 2 TMT mixtures in an unbalanced design, with 11 samples allocated to each mixture. 16 replicates were spread equally between the early and late WT and cKO conditions, resulting in four replicates per condition. Both the uninfected WT and cKO contained 3 replicates, with mixture one allocating one replicate to uninfected WT and two replicates to uninfected cKO. Conversely, mixture two contained one replicate of uninfected cKO and two uninfected WT. This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run. There was a 90\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled KO\_Uninfect, KO\_Early, KO\_Late, WT\_Uninfect, WT\_Early, and WT\_Late. 9 total comparisons were made, namely KO\_Early-WT\_Early, KO\_Late-WT\_Late, KO\_Uninfected-WT\_Uninfected, KO\_Early-KO\_Uninfected, KO\_Late-KO\_Uninfected, WT\_Early-WT\_Uninfected, WT\_Late-WT\_Uninfected, Infected-Uninfected, and KO-WT. Since the dataset was a biological investigation, the true positive modifications were unknown.

@@

\medskip \noindent {\bf Experimental Design:} Cunningham et al. \cite{Cunningham2015} investigated the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. The experiment profiled ubiquitination sites, and analyzed changes in the modified site abundance. The experiment had four conditions, CCCP, USP30 over expression (USP30 OE), Combo, and Control. Cell lines were used to create two biological replicates per condition. The abundance of modified peptides was quantified with label-free LC-MS/MS.

@@

Consider a label-free LC-MS/MS experiment in the special case of a balanced design with $I$ conditions and $J$ biological replicates (mass spectrometry runs) per condition.  For one protein, the PTM site is represented by $K$ spectral features (peptide ions, distinguished by their cleavage residues and charge states). The log-intensity (base 2) of Feature $k$, in Run $j$ of Condition $i$ is denoted by $y_{ijk}^{\ast}$. Conversely, the unmodified protein is represented by $L$ spectral features, and the log-intensity of Feature $l$ from the unmodified peptides in the same run is denoted by $y_{ijl}$. The features can be quantified as part of a same mass spectrometry run, or in separate enrichment and global proteome profiling runs.

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and the corresponding unmodified features of the protein in a special case of 2 conditions and 2 biological replicates per condition. The number of quantified features varies across replicate LC-MS/MS runs of the same sample, and across conditions. There may be feature outliers which could be caused by random variation or measurement error. Additionally, some spectral features can be missing. The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.

The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models. For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$. Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.

\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} = 0 \\
H_{a}: \Delta = \mu_{1}^{\ast} - \mu_{2}^{\ast} \neq 0
\end{aligned}
\label{eq:conv_null_hyp}
\end{equation}

However, these population parameters are inherently convoluted with changes in the unmodified protein. To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$. We estimate these quantities in the same way as for the PTM, using their corresponding features and observations, resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$. These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

@@

\noindent {\bf Two-sample $t$-test} Many investigations perform differential abundance analysis of PTMs using two-sample $t$-test or its extensions~\cite{Schwammle2015}\cite{THOMAS2020}. While simple, these approaches do not fully account for all sources of variations, and are not directly applicable to experiments with complex designs. These include experiments with comparisons of multiple conditions, acquisition in multiple batches, and unbalanced designs. Additionally, while these approaches can be applied to experiments studying PTMs, there is not a self contained, straightforward implementation of the methods, making application challenging.

When modeling the simple experiment seen in Figure \ref{fig:data-structure}, we are interested in answering the Null Hypothesis in Equation \ref{eq:null_hyp}. To estimate the parameters of interest in the trivial model, the approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another~\cite{Schwammle2015}. The abundance in each run is taken as input and is often estimated by sum of peak intensities. For example, the log-abundance estimate for the PTM in Run $j$ of Condition $i$ is given by Equation \ref{eq:feature_estimation}.
